---
title: "Drawing Plots"
author: "Peter Claussen"
date: "7/2/2020"
output:
  html_document: default
  pdf_document: default
bibliography: biblio.bib
---

```{r setup, include=FALSE}
library(ggplot2)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00", "#F0E442","#CC79A7","#000000","#734f80", "#2b5a74", "#004f39", "#787221", "#003959", "#6aaf00", "#663cd3")
add.metric <- function(data, origin=c(-1,-1)) {
  if(any(origin<0)) {
    origin[1] <- min(data$X)
    origin[2] <- min(data$Y)
  }
  data$Longitude <- data$X - origin[1]
  data$Latitude <- data$Y - origin[2]
  latMid <- (min(data$Y) + max(data$Y))/2
  m_per_deg_lat = 111132.954 - 559.822 * cos( 2.0 * latMid ) + 1.175 * cos( 4.0 * latMid)
  m_per_deg_lon = (3.14159265359/180 ) * 6367449 * cos ( latMid )
  data$Longitude <- data$Longitude*m_per_deg_lon
  data$Latitude <- data$Latitude*m_per_deg_lat
  return(data)
}
```

# Traditional Small-plot Trials

Much of the basic work on spatial analysis of crop yield data was based on homogeneity trials. Briefly, a homogeneity trial is an agronomic experiment without treatments. A field will be planted in discrete plots as is common practice in most field experiments, but all plots will be managed the same - same crop varieties, same fertilizers, same tillage, etc. 

The only variation among plots can than be assumed to be due to inherent variability in soil `fertility`. This gives a measure of simple spatial variance components. 

McCullagh and Clifford [-@mccullagh.p-2006] considered spatial variance components over several classical homogeneity trial data sets (among other trials). Their analysis is avialable at www.stat.uchicago.edu/wpmcc/reml. I'm include some of their code from `fairfield.R` here, and we'll use it to draw a plot map.

```{r}
metre <- 1.0
inch <- 2.54/100 * metre
foot <- 12 * inch
kilo <- 1.0
pound <- kilo / 2.204622
gram <- kilo/1000

rowwidth <- 0.5 * foot
colwidth <- 1.0 * foot
#PMC according to Fairfield, there were four rows (6 inches apart) 
# on each side and one foot on each end discarded at harvest
#rowsep <- 0 * foot
#colsep <- 0 * foot

rowsep <- 2 * foot
colsep <- 1 * foot
nrows <- 30
ncols <- 36
n <- nrows*ncols

y <- t(matrix(scan("./McCullagh/fairfield1.dat"), ncols, nrows))	# yield per plot
y <- matrix(y, n, 1) * 0.1  # to convert yield to grams
```

There a slicker way to do this in R, but I'm going to go a more manual but less error prone route

```{r}
fairfield.dat <- data.frame(Yield = rep(0,n),
                            Longitude = rep(0,n),
                            Latitude = rep(0,n))
idx <- 0
for (i in 1:nrows) {
  for(j in 1:ncols) {
    idx <- idx+1
    fairfield.dat$Latitude[idx]=i*(rowwidth+rowsep)
    fairfield.dat$Longitude[idx]=j*(colwidth+colsep)
    fairfield.dat$Yield[idx] <- y[idx]
  }
}
```


```{r}
fairfield.dat$e <- fairfield.dat$Longitude + rowwidth
fairfield.dat$w <- fairfield.dat$Longitude - rowwidth
fairfield.dat$n <- fairfield.dat$Latitude + colwidth
fairfield.dat$s <- fairfield.dat$Latitude - colwidth
```

```{r,fig.width=8,fig.height=6}
base.plot <- ggplot(fairfield.dat, aes(Longitude, Latitude)) + geom_point(aes(colour = Yield),size = 1) +
  scale_colour_gradient(low=cbPalette[7], high=cbPalette[4])

base.plot <- base.plot + geom_segment(aes(x = e, y = s, xend = w, yend = s), 
                                      data = fairfield.dat,color = cbPalette[3])
base.plot <- base.plot + geom_segment(aes(x = e, y = n, xend = w, yend = n), 
                                      data = fairfield.dat,color = cbPalette[3])
base.plot <- base.plot + geom_segment(aes(x = e, y = n, xend = e, yend = s), 
                                      data = fairfield.dat,color = cbPalette[3])
base.plot <- base.plot + geom_segment(aes(x = w, y = n, xend = w, yend = s), 
                                      data = fairfield.dat,color = cbPalette[3])

base.plot
```



# Yield Monitor Data

To illustrate the physical difference in the concept of *plot* as applied to yield monitor data, we will draw boundries associated with yield data points. We start with a subset of a typical yield data set, and change the names for convenience.
```{r}
load(file='../ManagementZoneML/home.squares.Rda')
harvest.dat <- home.2013.dat
# I have several data sets in this Rda, this line can be changed to quickly change the set.
#harvest.dat <- home.2015.dat - wheat, has drainage lines
#harvest.dat <- home.2016.dat # very uneven plot sizes
#harvest.dat <- home.2017.dat # quite homogeneous in part of the field, but some bunching
#harvest.dat <- home.2018.dat # similar to 2017

# The column names are taken from the raw data and are a bit obscure, give them more meaningful names.
names(harvest.dat)[4] <- 'Yield'
names(harvest.dat)[12] <- 'Longitude'
names(harvest.dat)[13] <- 'Latitude'

# to plot only a 80m by 40m square
#harvest.dat <- harvest.dat[harvest.dat$Longitude<80,]
#harvest.dat <- harvest.dat[harvest.dat$Latitude<40,]

# to get a similar number of plots (approximately),
plots <- length(fairfield.dat$Yield)
for(i in 1:200) {
  mask <- harvest.dat$Longitude<(i*8) & harvest.dat$Latitude<i
  if(sum(mask)>plots) {
    break
  }
}
harvest.dat <- harvest.dat[mask,]
```

We use variables from the yield monitor to determine plot dimensions. 

- DISTANCE : Distance (feet) Traveled From Previous Point
- SWATHWIDTH : Width in feet of a channel in the combine header. There are 7 channels in the data, and so we should be able to multiply this by 7, but as we see, this is not quite right. I'll have to go manually measure the width of the various headers used for harvesting different crops.

We will need to convert these to metric.
```{r}
harvest.dat$DISTANCE <- harvest.dat$DISTANCE*0.3048
harvest.dat$SWATHWIDTH <- harvest.dat$SWATHWIDTH*(harvest.dat$Swaths+2)*0.3048
```

We will draw out our plot area on the assumption that the `Yield` data point associated with `Longitute` and `Latitude` is the rectangle just behind that point. We use `DISTANCE` to determine to height (relative to direciton of travel) of the plot, and assume the yield GPS point is in the center of the `SWATH`. I've added a small offset to the trailing plot line prevent overlap, and will be plotting this in gray to help visualize direction of travel.

```{r}
harvest.dat$front <- harvest.dat$Latitude
harvest.dat$rear <- harvest.dat$Latitude - harvest.dat$DISTANCE + 0.1
harvest.dat$right <- harvest.dat$Longitude + harvest.dat$SWATHWIDTH/2
harvest.dat$left <- harvest.dat$Longitude - harvest.dat$SWATHWIDTH/2
```

But we also need to rotate the rectangle to oriented with the direction of travel. This will be simplest, I think, if we calculate an offset from front to back and from right to left, oriented on the dirction of travel


```{r}
sample.dat <- harvest.dat[harvest.dat$Longitude <25,]
sample.dat <- sample.dat[sample.dat$Latitude <8,]
```

```{r,fig.width=12,fig.height=4}
sample.dat$DistanceAngle <- sample.dat$Heading*pi/180
sample.dat$DistanceRise <- cos(sample.dat$DistanceAngle)*sample.dat$DISTANCE
sample.dat$DistanceRun <- sin(sample.dat$DistanceAngle)*sample.dat$DISTANCE

sample.dat$SwathAngle <- (sample.dat$Heading-90)*pi/180
sample.dat$SwathRise <- cos(sample.dat$SwathAngle)*sample.dat$SWATHWIDTH/2
sample.dat$SwathRun <- sin(sample.dat$SwathAngle)*sample.dat$SWATHWIDTH/2

base.plot <- ggplot(sample.dat, aes(Longitude, Latitude)) + geom_point(aes(colour = Yield),size = 3) +
  scale_colour_gradient(low=cbPalette[7], high=cbPalette[4])

#front line (left to right)
base.plot <- base.plot + geom_segment(aes(x = Longitude - SwathRun, 
                                          xend = Longitude + SwathRun,
                                          y = Latitude - SwathRise,
                                          yend = Latitude + SwathRise), 
                                      data = sample.dat,color = cbPalette[1],size = 1)
#rear line (left to right)
base.plot <- base.plot + geom_segment(aes(x = Longitude - SwathRun - DistanceRun, 
                                          xend = Longitude + SwathRun - DistanceRun,
                                          y = Latitude - DistanceRise - SwathRise,
                                          yend = Latitude - DistanceRise + SwathRise), 
                                      data = sample.dat,color = cbPalette[3],size = 1,linetype = "dotted")
#right line (front to back)
base.plot <- base.plot + geom_segment(aes(x = Longitude + SwathRun, 
                                          xend = Longitude + SwathRun - DistanceRun,
                                          y = Latitude + SwathRise,
                                          yend = Latitude + SwathRise - DistanceRise), 
                                      data = sample.dat,color = cbPalette[3],size = 1)
#left line (front to back)
base.plot <- base.plot + geom_segment(aes(x = Longitude - SwathRun, 
                                          xend = Longitude - SwathRun - DistanceRun,
                                          y = Latitude - SwathRise,
                                          yend = Latitude - SwathRise - DistanceRise), 
                                      data = sample.dat,color = cbPalette[3],size = 1)
base.plot
```


```{r,fig.width=10,fig.height=5,eval=FALSE}
base.plot <- ggplot(harvest.dat, aes(Longitude, Latitude)) + geom_point(aes(colour = Yield),size = 2) +
  scale_colour_gradient(low=cbPalette[7], high=cbPalette[4])

base.plot <- base.plot + geom_segment(aes(x = right, y = rear, xend = left, yend = rear), 
                                      data = harvest.dat,color = cbPalette[1])
base.plot <- base.plot + geom_segment(aes(x = right, y = front, xend = left, yend = front), 
                                      data = harvest.dat,color = cbPalette[2])
base.plot <- base.plot + geom_segment(aes(x = right, y = front, xend = right, yend = rear), 
                                      data = harvest.dat,color = cbPalette[2])
base.plot <- base.plot + geom_segment(aes(x = left, y = front, xend = left, yend = rear), 
                                      data = harvest.dat,color = cbPalette[2])
base.plot
```

\pagebreak

## Generalizing

Now that we've worked out a method, we'll package as a function; for this, this assumes a similar table structure as the file used above. We have been using this to subset the field to an approximate number of samples and to maintain a ratio. If we want an entire field, we set plots=0.


```{r}
plot.gps.grids <- function(harvest.dat,plots,ratio=1,residuals=FALSE) {
  if(plots>0) {
     for(i in 1:1000) {
      mask <- harvest.dat$Longitude<(i*ratio) & harvest.dat$Latitude<i
      if(sum(mask)>plots) {
        break
      }
    }
    harvest.dat <- harvest.dat[mask,]
  }

  harvest.dat$DISTANCE <- harvest.dat$DISTANCE*0.3048
  harvest.dat$SWATHWIDTH <- harvest.dat$SWATHWIDTH*(harvest.dat$Swaths+2)*0.3048
  
  #harvest.dat$front <- harvest.dat$Latitude
  #harvest.dat$rear <- harvest.dat$Latitude - harvest.dat$DISTANCE + 0.1
  #harvest.dat$right <- harvest.dat$Longitude + harvest.dat$SWATHWIDTH/2
  #harvest.dat$left <- harvest.dat$Longitude - harvest.dat$SWATHWIDTH/2
  
  low <- cbPalette[7]
  high <- cbPalette[4]
  if(residuals) {
    harvest.dat$Yield <- harvest.dat$Yield - mean(harvest.dat$Yield)
    high <- "#0072B2"
    low <- "#D55E00"
  }
harvest.dat$DistanceAngle <- harvest.dat$Heading*pi/180
harvest.dat$DistanceRise <- cos(harvest.dat$DistanceAngle)*harvest.dat$DISTANCE
harvest.dat$DistanceRun <- sin(harvest.dat$DistanceAngle)*harvest.dat$DISTANCE

harvest.dat$SwathAngle <- (harvest.dat$Heading-90)*pi/180
harvest.dat$SwathRise <- cos(harvest.dat$SwathAngle)*harvest.dat$SWATHWIDTH/2
harvest.dat$SwathRun <- sin(harvest.dat$SwathAngle)*harvest.dat$SWATHWIDTH/2

base.plot <- ggplot(harvest.dat, aes(Longitude, Latitude)) + geom_point(aes(colour = Yield),size = 2) +
  scale_colour_gradient(low=low, high=high)

#front line (left to right)
base.plot <- base.plot + geom_segment(aes(x = Longitude - SwathRun, 
                                          xend = Longitude + SwathRun,
                                          y = Latitude - SwathRise,
                                          yend = Latitude + SwathRise), 
                                      data = harvest.dat,color = cbPalette[1],size = .5)
#rear line (left to right)
base.plot <- base.plot + geom_segment(aes(x = Longitude - SwathRun - DistanceRun, 
                                          xend = Longitude + SwathRun - DistanceRun,
                                          y = Latitude - DistanceRise - SwathRise,
                                          yend = Latitude - DistanceRise + SwathRise), 
                                      data = harvest.dat,color = cbPalette[3],size = .3)
#right line (front to back)
base.plot <- base.plot + geom_segment(aes(x = Longitude + SwathRun, 
                                          xend = Longitude + SwathRun - DistanceRun,
                                          y = Latitude + SwathRise,
                                          yend = Latitude + SwathRise - DistanceRise), 
                                      data = harvest.dat,color = cbPalette[3],size = .3)
#left line (front to back)
base.plot <- base.plot + geom_segment(aes(x = Longitude - SwathRun, 
                                          xend = Longitude - SwathRun - DistanceRun,
                                          y = Latitude - SwathRise,
                                          yend = Latitude - SwathRise - DistanceRise), 
                                      data = harvest.dat,color = cbPalette[3],size = .3)
  return(base.plot)
}
```

\pagebreak

### 2013

```{r,fig.width=12,fig.height=4}
names(home.2013.dat)[4] <- 'Yield'
names(home.2013.dat)[12] <- 'Longitude'
names(home.2013.dat)[13] <- 'Latitude'
plot.gps.grids(home.2013.dat,plots,ratio=8,residuals=FALSE) 
plot.gps.grids(home.2013.dat,plots,ratio=8,residuals=TRUE) 
```


```{r,fig.width=16,fig.height=12}
plot.gps.grids(home.2013.dat,plots=0,ratio=8,residuals=FALSE) 
```

```{r,fig.width=24,fig.height=18}
soybean.2013.dat <- read.csv('../ManagementZoneML/yield/Home Soybeans 2013.csv')
soybean.2013.dat$Yield <- soybean.2013.dat$VRYIELDVOL
soybean.2013.dat <- add.metric(soybean.2013.dat)
plot.gps.grids(soybean.2013.dat,plots=0,residuals=FALSE) 
```


### 2015

```{r,fig.width=12,fig.height=4}
names(home.2015.dat)[4] <- 'Yield'
names(home.2015.dat)[12] <- 'Longitude'
names(home.2015.dat)[13] <- 'Latitude'
plot.gps.grids(home.2015.dat,plots,ratio=8,residuals=TRUE) 
```

```{r,fig.width=16,fig.height=12}
plot.gps.grids(home.2015.dat,plots=0,ratio=8,residuals=FALSE) 
```

```{r,fig.width=24,fig.height=18}
wheat.2015.dat <- read.csv('../ManagementZoneML/yield/Home Wheat 2015.csv')
wheat.2015.dat$Yield <- wheat.2015.dat$VRYIELDVOL
wheat.2015.dat <- add.metric(wheat.2015.dat)
plot.gps.grids(wheat.2015.dat,plots=0,residuals=FALSE) 
```

### 2016

```{r,fig.width=12,fig.height=4}
names(home.2016.dat)[4] <- 'Yield'
names(home.2016.dat)[12] <- 'Longitude'
names(home.2016.dat)[13] <- 'Latitude'
plot.gps.grids(home.2016.dat,plots,ratio=8,residuals=FALSE) 
```


```{r,fig.width=16,fig.height=12}
plot.gps.grids(home.2016.dat,plots=0,ratio=8,residuals=FALSE) 
```

```{r,fig.width=24,fig.height=18}
corn.2016.dat <- read.csv('../ManagementZoneML/yield/Home Corn 2016.csv')
corn.2016.dat$Yield <- corn.2016.dat$VRYIELDVOL
corn.2016.dat <- add.metric(corn.2016.dat)
plot.gps.grids(corn.2016.dat,plots=0,residuals=FALSE) 
```

### 2017

```{r,fig.width=12,fig.height=4}
names(home.2017.dat)[4] <- 'Yield'
names(home.2017.dat)[12] <- 'Longitude'
names(home.2017.dat)[13] <- 'Latitude'
plot.gps.grids(home.2017.dat,plots,ratio=8,residuals=FALSE) 
```

```{r,fig.width=16,fig.height=12}
plot.gps.grids(home.2017.dat,plots=0,ratio=8,residuals=FALSE) 
```

```{r,fig.width=24,fig.height=18}
soybean.2017.dat <- read.csv('../ManagementZoneML/yield/Home Soybeans 2017.csv')
soybean.2017.dat$Yield <- soybean.2017.dat$VRYIELDVOL
soybean.2017.dat <- add.metric(soybean.2017.dat)
plot.gps.grids(soybean.2017.dat,plots=0,residuals=FALSE) 
```

### 2018

```{r,fig.width=12,fig.height=4}
names(home.2018.dat)[4] <- 'Yield'
names(home.2018.dat)[12] <- 'Longitude'
names(home.2018.dat)[13] <- 'Latitude'
plot.gps.grids(home.2018.dat,plots,ratio=8,residuals=FALSE) 
```

```{r,fig.width=16,fig.height=12}
plot.gps.grids(home.2018.dat,plots=0,ratio=8,residuals=FALSE) 
```

```{r,fig.width=24,fig.height=18}
corn.2018.dat <- read.csv('../ManagementZoneML/yield/Home Corn 2018.csv')
corn.2018.dat$Yield <- rank(corn.2018.dat$VRYIELDVOL)
corn.2018.dat <- add.metric(corn.2018.dat)
plot.gps.grids(corn.2018.dat,plots=0,residuals=FALSE) 
```

# Variograms

## Fairfield

Reproducing the semivariogram from [@mccullagh.p-2006]

```{r}
row <- gl(nrows, 1, n)
col <- gl(ncols, nrows, n)
one <- matrix(rep(1,n), n, 1)
x1 <- as.numeric(row)*rowwidth
x2 <- as.numeric(col)*colwidth
dx1 <- x1 %*% t(one) - one %*% t(x1)
dx2 <- x2 %*% t(one) - one %*% t(x2)
dx <- sqrt(dx1^2 + dx2^2)
ydiff <- y %*% t(one) - one %*% t(y)
dx <- as.vector(dx)
pos <- (dx < 3) & (dx > 0)
dx <- dx[pos]
ydiff <- as.vector(ydiff)[pos]

plot(lowess(dx, ydiff^2, f=1/5), xlab="plot separation in m.",
	ylab="variance of yield difference", type="l")
title(main="smoothed variogram for Fairfield Smith wheat yield data")
```

Compare this with the `gstat` variogram

```{r}
library(gstat)
fairfield.var <- variogram(Yield~1, 
                          locations=~Longitude+Latitude,
                          data=fairfield.dat)
plot(fairfield.var)
```

The plot from `fairfield.R` is limited to a 3$m$ separation. This makes the graph look cleaner than it is. For comparison, the full variogram

```{r}
dx <- sqrt(dx1^2 + dx2^2)
ydiff <- y %*% t(one) - one %*% t(y)
dx <- as.vector(dx)
plot(lowess(dx, ydiff^2, f=1/5), xlab="plot separation in m.",
	ylab="variance of yield difference", type="l")

title(main="Variogram for Fairfield Smith wheat yield data")
```

And a summary of the raw data points - we use boxplot because there are too many points.

```{r}
dx.class <- as.factor(ceiling(dx))
plot(dx.class,as.vector(ydiff^2))
```

## Yield Monitor

### Full Field
```{r}
library(gstat)
names(home.2015.dat)[4] <- 'Yield'
names(home.2015.dat)[12] <- 'Longitude'
names(home.2015.dat)[13] <- 'Latitude'
home.2015.var <- variogram(Yield~1, 
                          locations=~Longitude+Latitude,
                          data=home.2015.dat)
plot(home.2015.var)
```

### Subset

```{r}
library(gstat)
harvest.var <- variogram(Yield~1, 
                          locations=~Longitude+Latitude,
                          data=harvest.dat)
plot(harvest.var)
harvest.vgm <- fit.variogram(harvest.var, vgm("Sph"))
harvest.vgm
plot(harvest.vgm,cutoff=100)
```

```{r}
n <- length(harvest.dat$Yield)
#row <- gl(nrows, 1, n)
#col <- gl(ncols, nrows, n)
one <- matrix(rep(1,n), n, 1)
x1 <- harvest.dat$Latitude
x2 <- harvest.dat$Longitude
y <-  harvest.dat$Yield
dx1 <- x1 %*% t(one) - one %*% t(x1)
dx2 <- x2 %*% t(one) - one %*% t(x2)
dx <- sqrt(dx1^2 + dx2^2)
ydiff <- y %*% t(one) - one %*% t(y)
dx <- as.vector(dx)
#pos <- (dx < 3) & (dx > 0)
#dx <- dx[pos]
#ydiff <- as.vector(ydiff)[pos]

plot(lowess(dx, ydiff^2, f=1/5), xlab="plot separation in m.",
	ylab="variance of yield difference", type="l")
title(main="Vvariogram for Home yield data")

```

# References

