---
title: "Plotting Response Surface"
author: "Peter Claussen"
date: "7/2/2020"
output:
  html_document: default
  pdf_document: default
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
library(gstat)
library(ggplot2)
grey <- "#999999"
orange <- "#E69F00"
skyblue <- "#56B4E9"
bluishgreen <- "#009E73"
yellow <- "#F0E442"
blue <- "#0072B2"
vermillion <- "#D55E00"
reddishpurple <- "#CC79A7"
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7",
                "#000000","#734f80", "#2b5a74", "#004f39", "#787221", "#003959", "#6aaf00", "#663cd3")
source('./R/add.metric.R')
source('./R/plot.gps.grids.R')
```


Most recent export from John Deere looks like it pools channels.

```{r,echo=FALSE}
corn.2013.dat <- read.csv('~/Work/git/statistics/test_cases/claussen/Stevens East/Stevens East 2013 Corn Harvest.csv')
corn.2013.dat$Yield <- rank(corn.2013.dat$VRYIELDVOL)
corn.2013.dat$Yield <- corn.2013.dat$Yield/max(corn.2013.dat$Yield)
corn.2013.dat <- add.metric(corn.2013.dat)
corn.2013.dat$Swaths <- 7
corn.2013.dat$SWATHWIDTH <- 5

soybean.2014.dat <- read.csv('~/Work/git/statistics/test_cases/claussen/Stevens East/Stevens East 2014 Soybeans Harvest.csv')
soybean.2014.dat$Yield <- rank(soybean.2014.dat$VRYIELDVOL)
soybean.2014.dat$Yield <- soybean.2014.dat$Yield/max(soybean.2014.dat$Yield)
soybean.2014.dat <- add.metric(soybean.2014.dat)
soybean.2014.dat$Swaths <- 7
soybean.2014.dat$SWATHWIDTH <- 5

corn.2015.dat <- read.csv('~/Work/git/statistics/test_cases/claussen/Stevens East/Stevens East 2015 Corn Harvest.csv')
corn.2015.dat$Yield <- rank(corn.2015.dat$VRYIELDVOL)
corn.2015.dat$Yield <- corn.2015.dat$Yield/max(corn.2015.dat$Yield)
corn.2015.dat <- add.metric(corn.2015.dat)
corn.2015.dat$Swaths <- 7
corn.2015.dat$SWATHWIDTH <- 5

soybean.2016.dat <- read.csv('~/Work/git/statistics/test_cases/claussen/Stevens East/Stevens East 2016 Soybeans Harvest.csv')
soybean.2016.dat$Yield <- rank(soybean.2016.dat$VRYIELDVOL)
soybean.2016.dat$Yield <- soybean.2016.dat$Yield/max(soybean.2016.dat$Yield)
soybean.2016.dat <- add.metric(soybean.2016.dat)
soybean.2016.dat$Swaths <- 7
soybean.2016.dat$SWATHWIDTH <- 5

corn.2017.dat <- read.csv('~/Work/git/statistics/test_cases/claussen/Stevens East/Stevens East 2017 Corn Harvest.csv')
corn.2017.dat$Yield <- rank(corn.2017.dat$VRYIELDVOL)
corn.2017.dat$Yield <- corn.2017.dat$Yield/max(corn.2017.dat$Yield)
corn.2017.dat <- add.metric(corn.2017.dat)
corn.2017.dat$Swaths <- 7
corn.2017.dat$SWATHWIDTH <- 5

soybean.2018.dat <- read.csv('~/Work/git/statistics/test_cases/claussen/Stevens East/Stevens East 2018 Soybeans Harvest.csv')
soybean.2018.dat$Yield <- rank(soybean.2018.dat$VRYIELDVOL)
soybean.2018.dat$Yield <- soybean.2018.dat$Yield/max(soybean.2018.dat$Yield)
soybean.2018.dat <- add.metric(soybean.2018.dat)
soybean.2018.dat$Swaths <- 7
soybean.2018.dat$SWATHWIDTH <- 5

corn.2013.dat$Year <- 2013
soybean.2014.dat$Year <- 2014
corn.2015.dat$Year <- 2015
soybean.2016.dat$Year <- 2016
corn.2017.dat$Year <- 2017
soybean.2018.dat$Year <- 2018

combined.dat <- rbind(corn.2013.dat,soybean.2014.dat,corn.2015.dat,soybean.2016.dat,corn.2017.dat,soybean.2018.dat)
```

```{r,fig.width=8,fig.height=8,echo=FALSE}
ggplot(combined.dat, aes(Longitude,Latitude)) + 
geom_point(aes(colour = Yield),size=1) + 
scale_colour_gradient2(low=vermillion, mid=yellow, high=blue,midpoint = 0.5) +
#scale_colour_gradient(low=cbPalette[8], high=cbPalette[5]) +
labs(colour = "Rank (normalized)", x="X (m)", y="Y (m)") + facet_wrap(~ Year)
```

### 2013

```{r,fig.width=6,fig.height=8}
plot.gps.grids(corn.2013.dat,plots=0,residuals=TRUE) 
```

### 2014

```{r,fig.width=6,fig.height=8}
plot.gps.grids(soybean.2014.dat,plots=0,residuals=TRUE) 
```


### 2015

```{r,fig.width=6,fig.height=8}
plot.gps.grids(corn.2015.dat,plots=0,residuals=TRUE) 
```

### 2016

```{r,fig.width=6,fig.height=8}
plot.gps.grids(soybean.2016.dat,plots=0,residuals=TRUE) 
```

### 2017

```{r,fig.width=6,fig.height=8}
plot.gps.grids(corn.2017.dat,plots=0,residuals=TRUE) 
```

### 2018


```{r,fig.width=6,fig.height=8}
plot.gps.grids(soybean.2018.dat,plots=0,residuals=TRUE) 
```




# Variograms

## Yield Monitor

## Functional models for Random Fields

### LOESS

```{r}
#when using VRYIELDRY
#spans <- c(0.05, 0.1, 0.2, 0.4, 0.6, 0.8)

#note - at 2^(-6) the edges are problematic
spans <- 2^(-5:0)

LOESS <- vector(length(spans),mode='list')
Variograms <- vector(length(spans),mode='list')
Residuals <- vector(length(spans),mode='list')
Estimates <- vector(length(spans),mode='list')
loess.dat <- NULL
idx <- 0
for(span in spans) {
  idx <- idx + 1
  tmp.dat <- corn.2015.dat[,c('Longitude','Latitude','Yield')]
  tmp.loess <- loess(Yield ~ Longitude + Latitude, span=span ,data=tmp.dat)
  tmp.dat$Yield <- predict(tmp.loess)
  tmp.dat$residuals <- residuals(tmp.loess)
  tmp.dat$Span = span
  loess.dat <- rbind(loess.dat, tmp.dat)
  Estimates[[idx]] <- tmp.dat
  tmp.var <- variogram(Yield~1, 
                          locations=~Longitude+Latitude,
                          alpha=c(0,45,90),
                          data=tmp.dat)
  Variograms[[idx]] <- tmp.var
  #plot(tmp.dat$VRYIELDVOL)
  tmp.var <- variogram(residuals~1, 
                          locations=~Longitude+Latitude,
                          alpha=c(0,45,90),
                          data=tmp.dat)
  Residuals[[idx]] <- tmp.var
  #plot(tmp.var,main=paste('residuals',span))
}
```



```{r,fig.width=7,fig.height=8}
#loess.dat <- rbind(loess.dat, tmp.dat)

ggplot(loess.dat, aes(Longitude,Latitude)) + 
geom_point(aes(colour = Yield),size=1) + 
scale_colour_gradient2(low=vermillion, mid=yellow, high=blue,midpoint = mean(loess.dat$Yield)) +
#scale_colour_gradient(low=cbPalette[8], high=cbPalette[5]) +
labs(colour = "LOESS", x="X (m)", y="Y (m)") + facet_wrap(~ Span)
```

```{r,fig.width=7,fig.height=8}
ggplot(loess.dat, aes(Longitude,Latitude)) + 
geom_point(aes(colour = residuals),size=1) + 
scale_colour_gradient2(low=vermillion, mid=yellow, high=blue,midpoint = 0) +
#scale_colour_gradient(low=cbPalette[8], high=cbPalette[5]) +
labs(colour = "Residuals", x="X (m)", y="Y (m)") + facet_wrap(~ Span)
```

```{r,fig.width=6,fig.height=3}
plot(Variograms[[1]],main=paste('LOESS',spans[1]))
plot(Variograms[[2]],main=paste('LOESS',spans[2]))
plot(Variograms[[3]],main=paste('LOESS',spans[3]))
plot(Variograms[[4]],main=paste('LOESS',spans[4]))
plot(Variograms[[5]],main=paste('LOESS',spans[5]))
plot(Variograms[[6]],main=paste('LOESS',spans[6]))
```

```{r,fig.width=6,fig.height=3}
plot(Residuals[[1]],main=paste('Residuals',spans[1]))
plot(Residuals[[2]],main=paste('Residuals',spans[2]))
plot(Residuals[[3]],main=paste('Residuals',spans[3]))
plot(Residuals[[4]],main=paste('Residuals',spans[4]))
plot(Residuals[[5]],main=paste('Residuals',spans[5]))
plot(Residuals[[6]],main=paste('Residuals',spans[6]))
```

```{r,fig.width=8,fig.height=6}
library(moments)
par(mfrow=c(2,3))
for(i in 1:6) {
  hist(Estimates[[i]]$Yield,
       main=paste('LOESS',spans[i]),
       xlab=paste("sk=",format(skewness(Estimates[[i]]$Yield),digits=4),
                  "ku = ",format(kurtosis(Estimates[[i]]$Yield),digits=4))
       )
}
```

```{r,fig.width=8,fig.height=6}
library(moments)
par(mfrow=c(2,3))
for(i in 1:6) {
  hist(Estimates[[i]]$residuals,
       main=paste('Residuals',spans[i]),
       xlab=paste("sk=",format(skewness(Estimates[[i]]$residuals),digits=4),
                  "ku = ",format(kurtosis(Estimates[[i]]$residuals),digits=4))
       )
}
```

Borrowed from 7.grid_cells.Rmd (Workshop)

```{r}
grid.dists <- as.matrix(dist(cbind(corn.2015.dat$Longitude,corn.2015.dat$Latitude)))
grid.dists <- 1/grid.dists
diag(grid.dists) <- 0
library(ape)

Moran.tbl <- data.frame(Source=rep(0,7),
                        Expected=rep(0,7),
                        MoranI.Y=rep(NA,7),
                        SD.Y=rep(NA,7),
                        p.Y=rep(NA,7),
                        MoranI.Res=rep(NA,7),
                        SD.Res=rep(NA,7),
                        p.Res=rep(NA,7))

yield.I <- Moran.I(corn.2015.dat$Yield, grid.dists)
Moran.tbl$Source[1] <- 0
Moran.tbl$MoranI.Y[1] <- yield.I$observed
Moran.tbl$Expected <- yield.I$expected
Moran.tbl$SD.Y[1] <- yield.I$sd
Moran.tbl$p.Y[1] <- yield.I$p.value
Moran.tbl$MoranI.Res[1] <- 0
Moran.tbl$p.Res[1] <- 1
for(i in 1:6) {
  yield.I <-Moran.I(Estimates[[i]]$Yield, grid.dists)
  res.I <-Moran.I(Estimates[[i]]$residuals, grid.dists)
  Moran.tbl$Source[i+1] <- spans[i]
  Moran.tbl$MoranI.Y[i+1] <- yield.I$observed
  Moran.tbl$SD.Y[i+1] <- yield.I$sd
  Moran.tbl$p.Y[i+1] <- yield.I$p.value
  Moran.tbl$MoranI.Res[i+1] <- res.I$observed
  Moran.tbl$SD.Res[i+1] <- res.I$sd
  Moran.tbl$p.Res[i+1] <- res.I$p.value
}
Moran.tbl
```

```{r}
par(mfrow=c(1,3))
plot(MoranI.Y ~ Source,data=Moran.tbl)
plot(MoranI.Res ~ Source,data=Moran.tbl)
plot(MoranI.Res ~ MoranI.Y,data=Moran.tbl)
```

```{r}
#fit<-lm(Yield ~ bs(Longitude + Latitude),data = corn.2015.dat)

Yield1.lm <- lm(Yield ~ poly(Longitude, Latitude, degree=5), data=corn.2015.dat)
```

The response surface library `rsm` has some useful plots to help us visualize the trend surface

```{r,fig.width=8,fig.height=4}
library(rsm)
par(mfrow=c(1,3))
image(Yield1.lm, Latitude ~ Longitude)
contour(Yield1.lm,Latitude ~ Longitude, image = TRUE)
persp(Yield1.lm, Latitude ~ Longitude, zlab = "Yield, Poly 5")
```


```{r}
#image(x=loess.dat$Longitude, y=loess.dat$Latitude, z=loess.dat$Yield)
#contour(x=loess.dat$Longitude, y=loess.dat$Latitude, z=loess.dat$Yield, image = TRUE)
```

### B-splines

Perperoglou [-@perperoglou-2019] note that the `bs` spline function in R may produce spurious results when used with 2-dimensional data.

### Natural Splines


### Generalized Additive Models

`gamlss`

#### `mgcv`
 Adapted from https://stackoverflow.com/questions/16554690/interpolate-in-a-3-dimensional-spline-in-r
```{r}
library(mgcv)
b1 <- gam(Yield ~ s(Longitude,Latitude),data=corn.2015.dat)
vis.gam(b1)
title("t.p.r.s")
corn.2015.dat$Spline <- predict(b1)
```

```{r,fig.width=8,fig.height=4}
par(mfrow=c(1,3))
image(b1, Latitude ~ Longitude)
contour(b1,Latitude ~ Longitude, image = TRUE)
persp(b1, Latitude ~ Longitude, zlab = "Yield, Spline")
```

```{r,fig.width=3,fig.height=4}
ggplot(corn.2015.dat, aes(Longitude,Latitude)) + 
geom_point(aes(colour = Spline),size=1) + 
scale_colour_gradient2(low=vermillion, mid=yellow, high=blue,midpoint = mean(corn.2015.dat$Spline)) +
#scale_colour_gradient(low=cbPalette[8], high=cbPalette[5]) +
labs(colour = "Residuals", x="X (m)", y="Y (m)")
```


## model 2 with te() smooths
```{r}
b2 <- gam(Yield ~ te(Longitude,Latitude),data=corn.2015.dat)
vis.gam(b2)
title("tensor product")
```

```{r,fig.width=8,fig.height=4}
par(mfrow=c(1,3))
image(b2, Latitude ~ Longitude)
contour(b2,Latitude ~ Longitude, image = TRUE)
persp(b2, Latitude ~ Longitude, zlab = "Yield, Poly 5")
```

## model 3 te() smooths specifying margin bases
```{r}
b3 <- gam(Yield ~ te(Longitude,Latitude,bs=c("tp", "tp")),data=corn.2015.dat)
vis.gam(b3)
title("tensor product")
```

```{r,fig.width=8,fig.height=4}
par(mfrow=c(1,3))
image(b3, Latitude ~ Longitude)
contour(b3,Latitude ~ Longitude, image = TRUE)
persp(b3, Latitude ~ Longitude, zlab = "Yield, TE")
```


```{r,fig.width=8,fig.height=4}
par(mfrow=c(1,3))
contour(b1,Latitude ~ Longitude, image = TRUE)
contour(b2,Latitude ~ Longitude, image = TRUE)
contour(b3,Latitude ~ Longitude, image = TRUE)
```

```{r}
gam1 <- gam(Yield ~ s(Longitude,Latitude),data=corn.2013.dat)
gam2 <- gam(Yield ~ s(Longitude,Latitude),data=corn.2015.dat)
gam3 <- gam(Yield ~ s(Longitude,Latitude),data=soybean.2016.dat)
gam4 <- gam(Yield ~ s(Longitude,Latitude),data=corn.2017.dat)
gam5 <- gam(Yield ~ s(Longitude,Latitude),data=soybean.2018.dat)
```

```{r,fig.width=8,fig.height=8}
par(mfrow=c(2,3))
contour(gam1,Latitude ~ Longitude, image = TRUE)
contour(gam2,Latitude ~ Longitude, image = TRUE)
contour(gam3,Latitude ~ Longitude, image = TRUE)
contour(gam4,Latitude ~ Longitude, image = TRUE)
contour(gam5,Latitude ~ Longitude, image = TRUE)
```


```{r,fig.width=8,fig.height=8}
par(mfrow=c(2,3))
image(gam1,Latitude ~ Longitude)
image(gam2,Latitude ~ Longitude)
image(gam3,Latitude ~ Longitude)
image(gam4,Latitude ~ Longitude)
image(gam5,Latitude ~ Longitude)
```

See https://www.r-bloggers.com/r-as-gis-for-ecologists/ for plots
https://www.r-bloggers.com/colored-3d-map/

# References

Seemingly unrelated regression

