---
title: "Plotting Response Surface"
author: "Peter Claussen"
date: "7/2/2020"
output:
  html_document: default
  pdf_document: default
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
library(gstat)
library(ggplot2)
grey <- "#999999"
orange <- "#E69F00"
skyblue <- "#56B4E9"
bluishgreen <- "#009E73"
yellow <- "#F0E442"
blue <- "#0072B2"
vermillion <- "#D55E00"
reddishpurple <- "#CC79A7"
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7",
                "#000000","#734f80", "#2b5a74", "#004f39", "#787221", "#003959", "#6aaf00", "#663cd3")

source('./R/remove.harvest.outliers.fn.R')
source('./R/spatial.model.selection.R')
source('./R/select.best.variogram.R')
source('./R/spatial.selection.plots.R')
source('./R/read.yield.data.R')
source('./R/combined.map.R')
source('./R/add.metric.R')
source('./R/plot.gps.grids.R')
```

# Introduction 

We have several years of yield data from a single, roughly rectangular field. We'll use this to develop a method for finding a response surface for yield for each year, then consider how to combine the different years into a single average yield map.

We have data from 2013-2021, but 2014 is incomplete and 2018 is missing. The data are stored as a part of the ManagementZoneML project, and have been anonymized to substitute GPS coordinates Easting and Northing to distance in meters from a common origin, denoted by Easting and Northing; the common coordinate is a lower-left corner. Fields have been grouped by geographic location, so the common origin may be shared among multiple fields over multiple years. See [`Anonymization.Rmd`](./ManagementZoneML/Anonymization.Rmd) for details.

I've written a utility function, `read.yield.data` that loads harvest data files and attaches attributes for `Year`, `Crop` and field identifier, a single character A,B,.... In this document, we focus on field `G`

```{r}
field = 'G'
root = '../ManagementZoneML/data/'
Years <- c(2013,2015,2016,2017,2018,2020,2021)
Crops <- c('Corn','Corn','Soybeans','Corn','Soybeans','Corn','Soybeans')

harvests <-vector(mode='list',length=length(Years))
for(i in 1:length(Years)) {
  harvests[[i]] <- read.yield.data(root,field,Years[i],Crops[i])
}
```

The `combined.map` function iterates over a list of individual fields and ranking yield within each year, then standardizing rank to a value from 0-1, denoted `Quantile`. This allows us to visualize the spatial trend in the field, independent of outliers.

```{r,fig.width=6,fig.height=8,echo=FALSE}
Maps6 <- combined.map(harvests)
ggplot(Maps6, aes(Easting,Northing)) + 
geom_point(aes(colour = Quantile),size=.5) + 
scale_colour_gradient2(low=vermillion, mid=yellow, high=blue, midpoint = 0.5) +
labs(colour = "Relative Yield Rank", x="Easting", y="Northing", title = "G, Multiple Years") + facet_wrap(~ Map)
```

# Variograms

## Yield Monitor

## Functional models for Random Fields

### GAM

We start by using generalized additive models to fit a smooth surface
```{r}
require(mgcv)
k.vec <- (1:6)*25

K <- vector(length(k.vec),mode='list')
Variograms <- vector(length(k.vec)+1,mode='list')
Residuals <- vector(length(k.vec)+1,mode='list')
Estimates <- vector(length(k.vec)+1,mode='list')
gam.dat <- NULL
idx <- 0

for(i in 0:length(k.vec)) {
  idx <- idx + 1
  tmp.dat <- harvests[[1]][,c('Northing','Easting','Yield')]
  if(i==0) {
    tmp.gam <- gam(Yield ~ 1 + s(Easting,Northing),data=tmp.dat)
  } else{
    fmla <- paste('Yield ~1 + s(Easting,Northing,k=',k.vec[i],')')
    tmp.gam <- gam(as.formula(fmla),data=tmp.dat)
  }

  tmp.dat$Yield <- predict(tmp.gam)
  tmp.dat$residuals <- residuals(tmp.gam)
  if(i==0) {
    tmp.dat$K = 0
  } else {
    tmp.dat$K = k.vec[i]
  }

  gam.dat <- rbind(gam.dat, tmp.dat)
  Estimates[[idx]] <- tmp.dat
  tmp.var <- variogram(Yield~1, 
                          locations=~Easting+Northing,
                          alpha=c(0,45,90),
                          data=tmp.dat)
  Variograms[[idx]] <- tmp.var
  #plot(tmp.dat$VRYIELDVOL)
  tmp.var <- variogram(residuals~1, 
                          locations=~Easting+Northing,
                          alpha=c(0,45,90),
                          data=tmp.dat)
  Residuals[[idx]] <- tmp.var
  #plot(tmp.var,main=paste('residuals',span))
}
```



```{r,fig.width=7,fig.height=8}
#loess.dat <- rbind(loess.dat, tmp.dat)

ggplot(gam.dat, aes(Easting,Northing)) + 
geom_point(aes(colour = Yield),size=1) + 
scale_colour_gradient2(low=vermillion, mid=yellow, high=blue,midpoint = mean(gam.dat$Yield)) +
#scale_colour_gradient(low=cbPalette[8], high=cbPalette[5]) +
labs(colour = "LOESS", x="X (m)", y="Y (m)") + facet_wrap(~ K)
```

```{r,fig.width=7,fig.height=8}
ggplot(gam.dat, aes(Easting,Northing)) + 
geom_point(aes(colour = residuals),size=1) + 
scale_colour_gradient2(low=vermillion, mid=yellow, high=blue,midpoint = 0) +
#scale_colour_gradient(low=cbPalette[8], high=cbPalette[5]) +
labs(colour = "Residuals", x="X (m)", y="Y (m)") + facet_wrap(~ K)
```

```{r,fig.width=6,fig.height=3}
plot(Variograms[[1]],main='GAM default')
plot(Variograms[[2]],main=paste('GAM',k.vec[1]))
plot(Variograms[[3]],main=paste('GAM',k.vec[2]))
plot(Variograms[[4]],main=paste('GAM',k.vec[3]))
plot(Variograms[[5]],main=paste('GAM',k.vec[4]))
plot(Variograms[[6]],main=paste('GAM',k.vec[5]))
plot(Variograms[[7]],main=paste('GAM',k.vec[6]))

```

```{r,fig.width=6,fig.height=3}
plot(Residuals[[1]],main='Residuals default')
plot(Residuals[[2]],main=paste('Residuals',k.vec[1]))
plot(Residuals[[3]],main=paste('Residuals',k.vec[2]))
plot(Residuals[[4]],main=paste('Residuals',k.vec[3]))
plot(Residuals[[5]],main=paste('Residuals',k.vec[4]))
plot(Residuals[[6]],main=paste('Residuals',k.vec[5]))
plot(Residuals[[7]],main=paste('Residuals',k.vec[6]))
```

```{r,fig.width=8,fig.height=6}
library(moments)
par(mfrow=c(2,3))
for(i in 1:7) {
  hist(Estimates[[i]]$Yield,
       main=paste('GAM'),
       xlab=paste("sk=",format(skewness(Estimates[[i]]$Yield),digits=4),
                  "ku = ",format(kurtosis(Estimates[[i]]$Yield),digits=4))
       )
}
```

```{r,fig.width=8,fig.height=6}
library(moments)
par(mfrow=c(2,3))
for(i in 1:7) {
  hist(Estimates[[i]]$residuals,
       main=paste('Residuals'),
       xlab=paste("sk=",format(skewness(Estimates[[i]]$residuals),digits=4),
                  "ku = ",format(kurtosis(Estimates[[i]]$residuals),digits=4))
       )
}
```

Borrowed from 7.grid_cells.Rmd (Workshop)

```{r}
grid.dists <- as.matrix(dist(cbind(harvests[[1]]$Easting,harvests[[1]]$Northing)))
grid.dists <- 1/grid.dists
diag(grid.dists) <- 0
library(ape)

Moran.tbl <- data.frame(Source=rep(0,7),
                        Expected=rep(0,7),
                        MoranI.Y=rep(NA,7),
                        SD.Y=rep(NA,7),
                        p.Y=rep(NA,7),
                        MoranI.Res=rep(NA,7),
                        SD.Res=rep(NA,7),
                        p.Res=rep(NA,7))

yield.I <- Moran.I(harvests[[1]]$Yield, grid.dists)
Moran.tbl$Source[1] <- 0
Moran.tbl$MoranI.Y[1] <- yield.I$observed
Moran.tbl$Expected <- yield.I$expected
Moran.tbl$SD.Y[1] <- yield.I$sd
Moran.tbl$p.Y[1] <- yield.I$p.value
Moran.tbl$MoranI.Res[1] <- 0
Moran.tbl$p.Res[1] <- 1
for(i in 1:6) {
  yield.I <-Moran.I(Estimates[[i]]$Yield, grid.dists)
  res.I <-Moran.I(Estimates[[i]]$residuals, grid.dists)
  Moran.tbl$Source[i+1] <- k.vec[i]
  Moran.tbl$MoranI.Y[i+1] <- yield.I$observed
  Moran.tbl$SD.Y[i+1] <- yield.I$sd
  Moran.tbl$p.Y[i+1] <- yield.I$p.value
  Moran.tbl$MoranI.Res[i+1] <- res.I$observed
  Moran.tbl$SD.Res[i+1] <- res.I$sd
  Moran.tbl$p.Res[i+1] <- res.I$p.value
}
Moran.tbl
```

```{r}
par(mfrow=c(1,3))
plot(MoranI.Y ~ Source,data=Moran.tbl)
plot(MoranI.Res ~ Source,data=Moran.tbl)
plot(MoranI.Res ~ MoranI.Y,data=Moran.tbl)
```

```{r}
#fit<-lm(Yield ~ bs(Easting + Northing),data = harvests[[1]])

Yield1.lm <- lm(Yield ~ poly(Easting, Northing, degree=5), data=harvests[[1]])
```

The response surface library `rsm` has some useful plots to help us visualize the trend surface

```{r,fig.width=8,fig.height=4}
library(rsm)
par(mfrow=c(1,3))
image(Yield1.lm, Northing ~ Easting)
contour(Yield1.lm,Northing ~ Easting, image = TRUE)
persp(Yield1.lm, Northing ~ Easting, zlab = "Yield, Poly 5")
```


```{r}
#image(x=loess.dat$Easting, y=loess.dat$Northing, z=loess.dat$Yield)
#contour(x=loess.dat$Easting, y=loess.dat$Northing, z=loess.dat$Yield, image = TRUE)
```

### B-splines

Perperoglou [-@perperoglou-2019] note that the `bs` spline function in R may produce spurious results when used with 2-dimensional data.

### Natural Splines


### Generalized Additive Models

`gamlss`

#### `mgcv`
 Adapted from https://stackoverflow.com/questions/16554690/interpolate-in-a-3-dimensional-spline-in-r
```{r}
library(mgcv)
b1 <- gam(Yield ~ s(Easting,Northing),data=harvests[[1]])
vis.gam(b1)
title("t.p.r.s")
harvests[[1]]$Spline <- predict(b1)
```

```{r,fig.width=8,fig.height=4}
par(mfrow=c(1,3))
image(b1, Northing ~ Easting)
contour(b1,Northing ~ Easting, image = TRUE)
persp(b1, Northing ~ Easting, zlab = "Yield, Spline")
```

```{r,fig.width=3,fig.height=4}
ggplot(harvests[[1]], aes(Easting,Northing)) + 
geom_point(aes(colour = Spline),size=1) + 
scale_colour_gradient2(low=vermillion, mid=yellow, high=blue,midpoint = mean(harvests[[1]]$Spline)) +
#scale_colour_gradient(low=cbPalette[8], high=cbPalette[5]) +
labs(colour = "Residuals", x="X (m)", y="Y (m)")
```


## model 2 with te() smooths
```{r}
b2 <- gam(Yield ~ te(Easting,Northing),data=harvests[[1]])
vis.gam(b2)
title("tensor product")
```

```{r,fig.width=8,fig.height=4}
par(mfrow=c(1,3))
image(b2, Northing ~ Easting)
contour(b2,Northing ~ Easting, image = TRUE)
persp(b2, Northing ~ Easting, zlab = "Yield, Poly 5")
```

## model 3 te() smooths specifying margin bases
```{r}
b3 <- gam(Yield ~ te(Easting,Northing,bs=c("tp", "tp")),data=harvests[[1]])
vis.gam(b3)
title("tensor product")
```

```{r,fig.width=8,fig.height=4}
par(mfrow=c(1,3))
image(b3, Northing ~ Easting)
contour(b3,Northing ~ Easting, image = TRUE)
persp(b3, Northing ~ Easting, zlab = "Yield, TE")
```


```{r,fig.width=8,fig.height=4}
par(mfrow=c(1,3))
contour(b1,Northing ~ Easting, image = TRUE)
contour(b2,Northing ~ Easting, image = TRUE)
contour(b3,Northing ~ Easting, image = TRUE)
```

```{r}
gam1 <- gam(Yield ~ s(Easting,Northing),data=harvests[[1]])
gam2 <- gam(Yield ~ s(Easting,Northing),data=harvests[[2]])
gam3 <- gam(Yield ~ s(Easting,Northing),data=harvests[[3]])
gam4 <- gam(Yield ~ s(Easting,Northing),data=harvests[[4]])
gam5 <- gam(Yield ~ s(Easting,Northing),data=harvests[[5]])
```

```{r,fig.width=8,fig.height=8}
par(mfrow=c(2,3))
contour(gam1,Northing ~ Easting, image = TRUE)
contour(gam2,Northing ~ Easting, image = TRUE)
contour(gam3,Northing ~ Easting, image = TRUE)
contour(gam4,Northing ~ Easting, image = TRUE)
contour(gam5,Northing ~ Easting, image = TRUE)
```


```{r,fig.width=8,fig.height=8}
par(mfrow=c(2,3))
image(gam1,Northing ~ Easting)
image(gam2,Northing ~ Easting)
image(gam3,Northing ~ Easting)
image(gam4,Northing ~ Easting)
image(gam5,Northing ~ Easting)
```

```{r}
summary(gam1)
summary(gam2)
summary(gam3)
summary(gam4)
summary(gam5)
```

See https://www.r-bloggers.com/r-as-gis-for-ecologists/ for plots
https://www.r-bloggers.com/colored-3d-map/

# References

Seemingly unrelated regression

