---
title: "Random Fields"
author: "Peter Claussen"
date: "7/2/2020"
output:
  html_document: default
  pdf_document: default
bibliography: biblio.bib
---

```{r setup, include=FALSE}
library(ggplot2)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00", "#F0E442","#CC79A7","#000000","#734f80", "#2b5a74", "#004f39", "#787221", "#003959", "#6aaf00", "#663cd3")
grey <- "#999999"
orange <- "#E69F00"
skyblue <- "#56B4E9"
bluishgreen <- "#009E73"
yellow <- "#F0E442"
blue <- "#0072B2"
vermillion <- "#D55E00"
reddishpurple <- "#CC79A7"
```

Random Fields


Abstract 

1 Introduction 
Suppose we have a uniformity trial, or a single season harvest. We write a model

$$
Y(\mathbf{s})=\mu(\mathbf{s})+Z(\mathbf{s})+e(\mathbf{s}) 
$$


where all variables are indexed by point $\mathbf{s}=\left\{s_1,\dots ,s_n\right\} \in \mathbb{R}^n$ (typically, $\mathbf{s}$ will be in $\mathbb{R}^2$ ). Y is the measured yield, usually in units such as bushels per acre. We take μ to be the expected yield and may be estimated by an average yield over multiple growing seasons, although for some fields we may have only the single season data. 
Both Z and e are random variables. Z is the deviation for a single growing season from the expected yield. e is a white-noise process associated with measurement error for a sample taken at point $\mathbf{s}$.
When we have single season data we will not be able to decompose μ and Z; so we may assume μ is constant
1. Or we can drop it from the model and include it in the distribution of Z.  ^1 and write
$Y(\mathbf{s})=\mu+Z(\mathbf{s})+e(\mathbf{s}) $
We will also require that Z be continuous, while e is not required to be continuous. That is, Z is a single realizaton of a random variable in $\mathbb{R}^2$, while e is an arbitrary (finite) set in $\mathbb{R}$.

2 Characteristics of a Random Field 
2.1 Stationarity 
Briefly, stationarty implies that the variance of the field Z is constant over space. When the field is non-stationary variance will be a function of \mathbf{s}. We might2. Might. In most cases, a field of this form is closed and bounded. Perhaps uniform vs absolute continuity?  ^2 state that when Z is stationary, then Z is uniformly continuous, while Z may be continuous, but not uniformly continuous, if non-stationary.
2.2 Isotropy 
We write the relationship (covariance) between yeild measured at two points $\mathbf{s},\mathbf{s}'$ as
$$
Cov(\mathbf{s},\mathbf{s}')=E[Y(\mathbf{s})-Y(\mathbf{s}')]=Z(\mathbf{s})-Z(\mathbf{s}')+e(\mathbf{s})-e(\mathbf{s}')=Z(\mathbf{s}-\mathbf{s}')+e(\mathbf{s}-\mathbf{s}') 
$$

When $Cov(\mathbf{s},\mathbf{s}')$ depends only on the distance between points, $\mathbf{s}-\mathbf{s}'$ a field is said to be isotropic. If covariance depends in direction3. Maybe denoted by $|\mathbf{s}| |\mathbf{s}'|cos\theta$ the the field is said to be anisotropic.

```{r}
yield.z.fn <- function(values,rng=3) {
  values <- values[!is.na(values)]
  yield.mean <- mean(values)
  yield.sd <- sd(values)
  return((values-yield.mean)/yield.sd)
}

load(file='../ManagementZoneML/home.squares.Rda')

home.2013.dat$Year <- 2013
home.2015.dat$Year <- 2015
home.2016.dat$Year <- 2016
home.2017.dat$Year <- 2017
home.2018.dat$Year <- 2018

home.2013.dat$Rank <- rank(home.2013.dat$VRYIELDVOL)
home.2015.dat$Rank <- rank(home.2015.dat$VRYIELDVOL)
home.2016.dat$Rank <- rank(home.2016.dat$VRYIELDVOL)
home.2017.dat$Rank <- rank(home.2017.dat$VRYIELDVOL)
home.2018.dat$Rank <- rank(home.2018.dat$VRYIELDVOL)

home.2013.dat$PercentRank <- 100*home.2013.dat$Rank/max(home.2013.dat$Rank)
home.2015.dat$PercentRank <- 100*home.2015.dat$Rank/max(home.2015.dat$Rank)
home.2016.dat$PercentRank <- 100*home.2016.dat$Rank/max(home.2016.dat$Rank)
home.2017.dat$PercentRank <- 100*home.2017.dat$Rank/max(home.2017.dat$Rank)
home.2018.dat$PercentRank <- 100*home.2018.dat$Rank/max(home.2018.dat$Rank)


home.2013.dat$Z <- yield.z.fn(home.2013.dat$VRYIELDVOL)
home.2015.dat$Z <- yield.z.fn(home.2015.dat$VRYIELDVOL)
home.2016.dat$Z <- yield.z.fn(home.2016.dat$VRYIELDVOL)
home.2017.dat$Z <- yield.z.fn(home.2017.dat$VRYIELDVOL)
home.2018.dat$Z <- yield.z.fn(home.2018.dat$VRYIELDVOL)

harvest.dat <- rbind(home.2013.dat,
  home.2015.dat,
  home.2016.dat,
  home.2017.dat,
  home.2018.dat)
# The column names are taken from the raw data and are a bit obscure, give them more meaningful names.
names(harvest.dat)[4] <- 'Yield'
names(harvest.dat)[12] <- 'Longitude'
names(harvest.dat)[13] <- 'Latitude'
```

```{r,fig.width=9,fig.height=5,echo=FALSE}
ggplot(harvest.dat, aes(Longitude,Latitude)) + 
geom_point(aes(colour = Yield),size=.5) + 
scale_colour_gradient(low=yellow, high=bluishgreen) +
labs(colour = "Yield", x="X (m)", y="Y (m)") + facet_wrap(~ Year)
```

```{r,fig.width=9,fig.height=5,echo=FALSE}
ggplot(harvest.dat, aes(Longitude,Latitude)) + 
geom_point(aes(colour = PercentRank),size=.5) + 
scale_colour_gradient(low=grey, high=orange) +
labs(colour = "Rank", x="X (m)", y="Y (m)") + facet_wrap(~ Year)
```

```{r,fig.width=9,fig.height=5,echo=FALSE}
ggplot(harvest.dat, aes(Longitude,Latitude)) + 
geom_point(aes(colour = PercentRank),size=.5) + 
scale_colour_gradient2(low=vermillion, mid=yellow, high=blue,midpoint = 50) +
labs(colour = "Rank", x="X (m)", y="Y (m)") + facet_wrap(~ Year)
```


```{r,fig.width=9,fig.height=5,echo=FALSE}
ggplot(harvest.dat, aes(Longitude,Latitude)) + 
geom_point(aes(colour = Z),size=.5) + 
scale_colour_gradient2(low=vermillion, mid=yellow, high=blue) +
labs(colour = "Z", x="X (m)", y="Y (m)") + facet_wrap(~ Year)
```


# References

